Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-07 arxiv-version/main.tex
minor
Split sections in separate files

% Math symbols and commands
\newcommand{\R}{\mathbb{R}}  % set of real numbers
\newcommand{\indicator}[1]{\mathbf{1}\left[#1 \right]} % indicator
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle} % inner product
\newcommand{\norm}[1]{\left\| #1 \right\|}  % norm of a vector or a matrix
%\DeclareMathOperator*{\deg}{deg}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\Exp}{\mathbf{E}}  % expected value
\DeclareMathOperator{\diff}{d \!} % differential
\DeclareMathOperator*{\polylog}{polylog}

@@

\section{Introduction}
\label{section:introduction}

We study the problem of \textsc{Online Multiclass Linear
Classification with Bandit Feedback}~\cite{Kakade-Shalev-Shwartz-Tewari-2008}.
The problem can be viewed as a game between the learner and the environment:
at each timestep $t$, the environment chooses an example $(x_t, y_t)$, where its feature
part $x_t \in \R^d$ is revealed to the learner.
\footnote{All our results apply to also to a simpler model of
the so called \emph{oblivious adversary} where the whole sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is chosen before the first
round of the game.}
Upon receiving this example, the
learner is asked to make a prediction $\hat{y}_t$.
the learner receives feedback immediately after making its prediction.
In contrast with the standard full information setting (where the feedback given is the correct label $y_t$), the feedback given is only a binary indicator
whether the prediction was correct or not. The protocol of the problem is formally stated below.

%In this problem, in each round, the
%learner receives a feature vector and is asked to predict one of $K$
%possible classes. To allow
%for meaningful use of randomization, we assume that the correct label in a
%particular round is chosen prior to algorithm's prediction in that
%round.
%Immediately after the learner makes its prediction, it
%receives feedback whether the prediction was correct or not. However, in
%contrast with the standard, so called full information setting, the algorithm
%does \emph{not} receive the correct label. The feedback is only binary.

\begin{algorithm}[h]
\caption{\textsc{Online Multiclass Classification with Bandit Feedback}
\label{algorithm:game-protocol}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Adversary \emph{secretly} chooses correct class label $y_t \in \{1,2,\dots,K\}$}
\chicheng{Or we simply say: Adversary chooses example $(x_t, y_t)$, where $x_t$ is reveal to the learner.}
\STATE{Predict class label $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \begin{cases}
0 & \text{if $\widehat y_t = y_t$,} \\
1 & \text{if $\widehat y_t \neq y_t$.}
\end{cases}$
}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

The performance is of the learner is measured by its cumulative number of mistakes
$\sum_{t=1}^T z_t = \sum_{t=1}^T \indicator{\widehat y_t \neq y_t}$. The bandit feedback
 is a natural feedback model motivated by applications including
online advertising and online content optimization, where the learner only gets to
see the loss of its action taken.
%where the binary feedback corresponds to a click
%The
%corresponding problem with full information feedback is solved by the classical
%\textsc{Multiclass Perceptron} algorithm and its variants (see e.g.
%\cite{Crammer-Dekel-Keshet-Shalev-Shwartz-Singer-2006}).

In this paper, we focus on the special case when the examples lie in $\R^d$ and
are \emph{linearly separable with a margin}.
\chicheng{I wonder if we should introduce the two notions of linear separability here,
because strong linear separability is also worth mentioning in the full-information setting.
}

In the full-information feedback
setting, it is well known that if the examples have norm at most $R$ and are
linearly separable with a margin $\gamma$ then the \textsc{Multiclass
Perceptron} algorithm makes at most $\lfloor 2(R/\gamma)^2 \rfloor$ mistakes.
This result is very satisfying since the upper bound on the number of mistakes
is information-theoretically optimal and at the same time the \textsc{Multiclass
Perceptron} algorithm has low time and memory complexity.

The bandit feedback setting, however, is much more challenging. To understand
the differences, we dissect the notion of linear separability in the multiclass
setting. Beside the standard notion of linear separability, we define also
\emph{strong linear separability}. For binary classification, these two
definitions are identical. However, they differ for multiclass classification.
For multiclass classification with $K$ classes the standard notion of linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside of the class lie in the complement of
the intersection of the halfspaces. Strong linear separability means that
examples from each class are separated from the remaining examples by a
\emph{single} hyperplane. We state the definitions precisely in
Section~\ref{section:notions-of-linear-separability}.

For the case when the examples are strongly linearly separable, we design a
simple and efficient algorithm with expected number of mistakes at most $(K-1)
\lfloor (R/\gamma)^2 \rfloor$. The algorithm can be viewed as running $K$ copies
of the \textsc{Binary Perceptron} algorithm, one copy for each class.
Intuitively speaking, the factor $K-1$ is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We also
prove that any (possibly randomized) algorithm makes $\Omega(K (R/\gamma)^2))$
mistakes in the worst case.

For the standard notion of linear separability, we propose several algorithms
with various time-complexity-versus-mistake-bound trade-offs. First class of
algorithms map the examples into a high-dimensional inner product space. In this
space the examples become \emph{strongly} linearly separable. This allows us to
run the algorithm for the strongly linearly separable case. All of these
algorithms can be kernelized. (Some of them have to be kernelized since the
underlying inner product space is infinite dimensional.) Kernelization does not
change the input-output behavior of the algorithm, it only affects its time
and memory complexity.

These algorithms differ by choice of the mapping or, equivalently, choice of the
kernel. The crux of the problem is the analysis of the approximation properties
of these mappings/kernels and the way margin in the original space $\R^d$ under
standard linear separability notion gets transformed into a margin in the
high-dimensional space under the strong separability notion. This problem is
related to the problem of learning intersection of halfspaces and has been
studied previously by \cite{Klivans-Servedio-2008}. As an important technical
achievement, we improve on the results of \cite{Klivans-Servedio-2008} by
removing completely the dependency on the original dimension $d$.

Finally, we propose a very different algorithm that is based on the obvious idea
that two points that are close enough must have the same label. Here, close
enough means that they cannot be separated from each other by a linear separator
with a margin.

All the algorithms run in time that is polynomial in the dimension of the
feature vectors, the number of classes and the number of rounds. At the same
time, all the algorithms make number of mistakes that does \emph{not} depend
on the number of rounds. Instead, their number of mistakes depends only on the
margin, number of classes, dimension of the feature vectors, and the radius of
the ball in which the data lie.

Finally, we study two questions related to the computational and
information-theoretic hardness of the problem. Any algorithm for the bandit
setting collects information in the form of so called \emph{strongly labeled}
and \emph{weakly labeled} examples. Strongly labeled examples are those for
which we know the class label. Weakly labeled example is an example for which we
know that class label can be anything except for a particular one class. First,
we show the offline problem of finding a linear multiclass separator consistent
with a set of strongly and weakly labeled examples is NP-hard. This result is a
consequence of the fact that finding an intersection of two halfspaces
consistent with a set of binary-labeled examples is
NP-hard~\citep{Blum-Rivest-1993}. Second, we show that any algorithm that uses
only strongly-labeled examples and ignores weakly labeled examples makes at
least $\Omega(???)$ mistakes.

\section{Related work}
\label{section:related-work}

\begin{itemize}
\item \cite{Abernethy-Rakhlin-2009}

\item \cite{Chen-Chen-Zhang-Chen-Zhang-2009}

\item \cite{Hazan-Kale-2011}

\item \cite{Beygelzimer-Orabona-Zhang-2017}

\item \cite{Foster-Kale-Luo-Mohri-Sridharan-2018}

\item \cite{Foster-Krishnamurthy-2018}
\end{itemize}

TODO

\section{Notions of linear separability}
\label{section:notions-of-linear-separability}

We define two notions of linear separability for multiclass classification. The
first notion is the standard notion of linear separability used in the proof of
the mistake bound for the \textsc{Multiclass Perceptron} algorithm. The second
notion is stronger, i.e. more restrictive. However, it is more suitable for the
bandit setting, since it allows for a simple and efficient algorithm; see
Section~\ref{section:algorithm-for-strongly-linearly-separable-data}.

\begin{definition}[Linear separability]
Let $(V,\ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, and let $\gamma$ be a positive real number. Let $(x_1, y_1), (x_2,
y_2), \dots, (x_T, y_T)$ be labeled examples in $V \times \{1,2,\dots,K\}$.

The examples are said to be \emph{linearly separable with a
margin $\gamma$} if there exist vectors $w_1, w_2, \dots, w_K \in V$ such
that
\begin{align}
\label{equation:linear-separability-1}
\sum_{i=1}^K \norm{w_i}^2 & \le 1 \; , \\
\label{equation:linear-separability-2}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \qquad \qquad \ip{x_t}{w_{y_t}} & \ge \ip{x_t}{w_i} + \gamma \; .
\end{align}

The examples are said to be \emph{strongly linearly separable with a
margin $\gamma$} if there exist vectors $w_1, w_2, \dots, w_K \in V$ such
that
\begin{align}
\label{equation:strong-linear-separability-1}
\sum_{i=1}^K \norm{w_i}^2 & \le 1 \; , \\
\label{equation:strong-linear-separability-2}
\forall t \in \{1,2,\dots,T\} \qquad \qquad \ip{x_t}{w_{y_t}} &\ge \gamma \; , \\
\label{equation:strong-linear-separability-3}
\forall t \in \{1,2,\dots,T\} \qquad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \qquad \qquad \ip{x_t}{w_i} & \le - \gamma \; .
\end{align}
\end{definition}

The notion of linear separability with a margin is standard. It used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. For completeness, we state the
algorithm in Appendix~\ref{section:multiclass-perceptron-proofs}. It is a
folklore result that if a set of labeled examples is separable with a margin
$\gamma$ and the norm of the examples is bounded by $R$ then \textsc{Multiclass
Perceptron} algorithm makes at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
mistakes. Another folklore result is that \textsc{Multiclass Perceptron} is
essentially optimal in the sense that any deterministic algorithm must make
$\left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the worst case. For
completeness, we give proofs of both of these results in
Appendix~\ref{section:multiclass-perceptron-proofs}.

The notion of strong linear separability is new.

It is easy to see that if labeled examples are strongly linearly
separable with margin $\gamma$ then they are linearly separable with margin
$2\gamma$. Indeed, if $w_1, w_2, \dots, w_K \in V$
satisfy \eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3} then they satisfy
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} with margin $\gamma' = 2\gamma$.

Similarly, if $K=2$ and labeled examples are linearly separable with a margin
$\gamma$ then they are strongly linearly separable with margin $\gamma/2$.
Indeed, if $w_1, w_2$ satisfy \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} then $w_1' = \frac{w_1 - w_2}{2}$, $w_2' =
\frac{w_2 - w_1}{2}$ and $\gamma'=\gamma/2$ satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3}. The last two conditions are easy
to see. The first condition follows from $\norm{w_i'}^2 \le (\frac{1}{2}
\norm{w_1} + \frac{1}{2} \norm{w_2})^2 \le \frac{1}{2}\norm{w_1}^2 +
\frac{1}{2}\norm{w_2}^2 \le \frac{1}{2}$ for $i=1,2$.

However, for any $K \ge 3$ and any inner product space of dimension at least
$2$, there exists a set of labeled examples that is linearly separable with a
positive margin $\gamma$ but it is not strongly linearly separable with any
positive margin $\gamma$. An example of such set of labeled examples is shown in
Figure~\ref{figure:linearly-separable-examples-with-margin}.

\begin{figure}
\begin{center}
\input{figures/linearly-separable-examples-with-margin}
\end{center}
\caption[]{The figure shows labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. In other words, each class lies in an intersection of two
halfspaces.

While the examples are linearly separable with a positive margin $\gamma$, they
are \emph{not} strongly linearly separable with any positive margin $\gamma$.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

\section{Algorithm for strongly linearly separable data}
\label{section:algorithm-for-strongly-linearly-separable-data}

In this section, we present an algorithm for \textsc{Online Multiclass
Classification with Bandit Feedback} and prove an upper bound on the number of
mistakes under the assumption that the examples are strongly linearly separable
with a margin. The algorithm is stated as
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} and
the mistake upper bound is stated as
\autoref{theorem:strongly-separable-example-mistake-upper-bound}.

The idea behind the algorithm is to use $K$ copies of the \textsc{Binary
Perceptron} algorithm. Each copy corresponds to a class. In each round, each
copy makes a binary decision whether the feature vector belongs its respective
class or not. If at least one copy makes a positive prediction, the algorithm
predicts accordingly. If multiple copies make positive predictions, the
algorithm chooses one of them arbitrarily. If all copies make negative
predictions, the algorithms makes a uniformly random guess.

\begin{algorithm}[h]
\caption{\textsc{Bandit Algorithm for Strongly Linearly Separable Examples}
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$. Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\ENDIF
\ELSE
\STATE{Predict $\widehat y_t \in S_t$ chosen arbitrarily}
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake upper bound]
\label{theorem:strongly-separable-example-mistake-upper-bound}
Let $(V, \ip{\cdot}{\cdot})$ be any inner product space, let $K$ be a positive
integer, let $\gamma$ a positive real number, let $R$ be a non-negative real number. If
$(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled examples in
$V \times \{1,2,\dots,K\}$ such that the examples are strongly separable with
margin $\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ then the
expected number of mistakes
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
makes is at most $(K-1) \lfloor (R/\gamma)^2 \rfloor$.
\end{theorem}

\begin{proof}
Let $M = \sum_{t=1}^T z_t$ be the number of mistakes the algorithm makes. Let $A
= \sum_{t ~:~ S_t \neq \emptyset} z_t$ be the number of mistakes in the rounds
when $S_t \neq \emptyset$ and let $B = \sum_{t ~:~ S_t = \emptyset} z_t$ be the
number of mistakes in the rounds when $S_t = \emptyset$. Obviously, $M = A + B$.
We upper bound $A$ and $B$ separately.

Let $C$ be the number of times line 12 gets executed. Let $U$ be the number of
times line 12 and line 19 get executed. In other words, $U$ is the number of
times the $K$-tuple of vectors $(w_1^{(t)}, w_2^{(t)}, \dots, w_K^{(t)})$ gets
updated. Clearly, $U = A + C$.

The key observation is that $\Exp[B] = (K-1) \Exp[C]$. The equality holds
because if $S_t = \emptyset$, there is $1/K$ probability that the algorithm
guesses the correct label and with probability $(K-1)/K$ algorithm's guess is
incorrect. Putting all the information together, we get that
\begin{align*}
\Exp[M]
& = \Exp[A] + \Exp[B] \\
& = \Exp[A] + (K-1) \Exp[C]  \\
& \le (K-1) \Exp[A + C] \\
& = (K-1) \Exp[U]  \; .
\end{align*}

To finish the proof, we need to upper bound the number of updates, $U$. We claim
that $U \le \lfloor (R/\gamma)^2 \rfloor$. The proof of this upper bound is
similar to the proof of the mistake bound for \textsc{Multiclass Perceptron}
algorithm. Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors that satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}.
The $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots, w_K^{(t)})$
changes only if there is an update in round $t$.
We investigate how $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ and
$\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$ change. If there is an update in round $t$,
we have
\begingroup
\allowdisplaybreaks
\begin{align*}
\sum_{i=1}^K \norm{w_i^{(t+1)}}^2
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t+1)}}^2 \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t}^2 \\
& = \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 + \underbrace{(-1)^{z_t} 2 \ip{w_{\widehat y_t}^{(t)}}{x_t}}_{\le 0} \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + R^2 \; .
\end{align*}
\endgroup
Therefore, after $U$ updates,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
$$
Similarly, if there is an update in round $t$, we have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + (-1)^{z_t} \ip{w_{\widehat y_t}^*}{x_t} \\
& \ge \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma
\end{align*}
where the last inequality follows from \eqref{equation:strong-linear-separability-2}
and \eqref{equation:strong-linear-separability-3} and analysis of the cases $z_t = 0$ and $z_t = 1$.
Thus, after $U$ updates,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma U \; .
$$
Applying Cauchy-Schwartz's inequality twice and assumption
\eqref{equation:strong-linear-separability-1}, we get that
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
\le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}}
\le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2}
\le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
$$
Combining the upper and lower bounds on $\sum_{i=1}^K \norm{w_i^{(T+1)}}^2$ inequalities, we get
$$
(\gamma U)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
$$
We conclude that $U \le (R/\gamma)^2$. Since $U$ is an integer, $U \le \lfloor (R/\gamma)^2 \rfloor$.
\end{proof}

The upper bound $(K-1) \lfloor (R/\gamma)^2 \rfloor$ on the expected number of
mistakes of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
optimal to within constant factor as long as the number of classes $K$ is
at most $O((R/\gamma)^2)$. This is proved in
\autoref{theorem:strongly-separable-example-mistake-lower-bound} below.

Irrespective of any conditions on $K$, $R$, and $\gamma$, a trivial lower bound
on the expected number of mistakes of any randomized algorithm is
$\frac{K-1}{2}$. (Adversary chooses the same example over and over.) However, it
is unclear if the trivial lower bound is the best possible if $K$ is large,
i.e., $\omega((R/\gamma)^2)$. We leave this as an open problem.

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-example-mistake-lower-bound}
Let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. For any (possibly randomized) algorithm for the \textsc{Online
Multiclass Classification with Bandit Feedback} with $K \le
\frac{1}{4}(R/\gamma)^2$ classes there exists an inner product space $(V,
\ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ in $V \times
\{1,2,\dots,K\}$ that are strongly linearly separable with margin $\gamma$, the
norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the expected
number of mistakes the algorithm makes is at least $\frac{K-1}{2} \left\lfloor
\frac{1}{16} (R/\gamma)^2 \right\rfloor$.
\end{theorem}

\begin{proof}
We use probabilistic method. Let $M = \left\lfloor \frac{1}{16} (R/\gamma)^2 \right\rfloor$.
Let $V = \R^{M+1}$ equipped with the standard inner product.  Let $e_1, e_2,
\dots, e_{M+1}$ be the standard orthonormal basis of $V$. We define
vectors $v_1, v_2, \dots, v_M \in V$ where $v_i = \frac{R}{\sqrt{2}}(e_i +
e_{M+1})$ for $i=1,2,\dots,M$. Let $\ell_1, \ell_2, \dots, \ell_M$ be chosen
i.i.d. uniformly at random from $\{1,2,\dots,K\}$ and independently of any
randomness used the by the algorithm. Let $T = M K$. We define examples $(x_1,
y_1), (x_2, y_2), \dots, (x_T, y_T)$ as follows:
$$
\begin{array}{lclclclcl}
(x_1, y_1) & = & (x_2, y_2) & = & \cdots & = & (x_K, y_K) & = & (v_1, \ell_1) \; , \\
(x_{K+1},y_{K+1}) & = & (x_{K+2}, y_{K+2}) & = & \cdots & = & (x_{2K}, y_{2K}) & = & (v_2, \ell_2) \; , \\
& \vdots & \\
(x_{(M-1)K+1}, y_{(M-1)K+1})  & = & (x_{(M-1)K+2}, y_{(M-1)K+2}) & = & \cdots & = & (x_{MK}, y_{MK}) & = & (v_M, \ell_M) \; .
\end{array}
$$

With probability one, norm of each example is $R$. We show that with probability
one, the examples are strongly separable with margin $\gamma$. In order to see
that consider $w_1^*, w_2^*, \dots, w_K^* \in V$ defined by
$$
w_i^* = 2 \sqrt{2} \frac{\gamma}{R} \left( \sum_{j ~:~ \ell_j = i} e_j \right) - \sqrt{2} \frac{\gamma}{R} e_{M+1} \qquad \quad \text{for $i=1,2,\dots,K$.}
$$
Since for any $i \in \{1,2,\dots,K\}$ and any $j \in \{1,2,\dots,M\}$,
$$
\ip{w_i^*}{v_j} =
\begin{cases}
\gamma & \text{if $\ell_j = i$,} \\
- \gamma & \text{if $\ell_j \neq i$,}
\end{cases}
$$
it means that $w_1^*, w_2^*, \dots, w_K^*$ satisfy
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}. Condition \eqref{equation:strong-linear-separability-1}
is satisfied since
$$
\sum_{i=1}^K \norm{w_i^*}^2
= 8 \frac{\gamma^2}{R^2} \sum_{j=1}^M \norm{e_j}^2 + 2 \frac{\gamma^2}{R^2} K \norm{e_{M+1}}^2
= 8 \frac{\gamma^2}{R^2} M + 2 \frac{\gamma^2}{R^2} K
\le \frac{1}{2} + \frac{1}{2}
= 1 \; .
$$

It remains to lower bound the expected number of mistakes of the algorithm. For
any $j \in \{1,2,\dots,M\}$, consider the expected number of mistakes the
algorithm makes in rounds $K(j-1) + 1, K(j-1) + 2, \dots, Kj$. In these rounds,
the algorithm is guessing the label $\ell_j$. Since $\ell_j$ is chosen uniformly
at random from the set $\{1,2,\dots,K\}$ and feedback is only binary, the
expected number of mistakes the algorithm makes in these rounds is at least
$\frac{K-1}{2}$. Altogether, the algorithm makes at least $\frac{K-1}{2} M$
mistakes in expectation.

Finally, since strong separability and norm condition hold with probability one,
there exists a particular (i.e. deterministic) sequence of examples for which
the algorithm makes at least $\frac{K-1}{2} M$ mistakes in expectation
over its internal randomization.
\end{proof}

\section{From linear separability to strong linear separability}
\label{section:from-linear-separability-to-strong-linear-separability}

In this section, we show how to construct a mapping $\phi$ from the unit ball of
$\R^d$ into a high dimensional inner product space that has the property that if
a set of labeled examples in the unit ball is linearly separable with a margin
$\gamma$, applying the mapping $\phi$ makes the examples \emph{strongly}
linearly separable with a margin $\gamma'$ and their norms are bounded by $R'$.
The parameters $\gamma'$ and $R'$ are functions of the old margin $\gamma$ and
the number of classes $K$, and the are specified in the theorems below.

Equipped with the mapping $\phi$, we can utilize
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} and
\autoref{theorem:strongly-separable-example-mistake-upper-bound} from the
previous section and we obtain an algorithm for linearly separable examples and
an upper bound on its number of mistakes. As a computational speed up, instead
of working with the mapped examples $\phi(x_1), \phi(x_2), \dots$ explicitly, we
can use the kernelized version of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples}
that uses a kernel function $k(x,x') = \ip{\phi(x)}{\phi(x')}$.

We construct several different mappings $\phi$. The mappings differ in the
parameters $R'$, $\gamma'$, time complexity of evaluating $\phi(x)$, and time
complexity of evaluating $k(x,x')$. Some of the mappings depend the original
margin $\gamma$. In practice, the margin parameter $\gamma$ is not known, which
makes these mappings impractical. However, one of the mappings we construct
does \emph{not} depend on $\gamma$.

The idea behind all the mappings is polynomial approximation. According to the
well known Stone-Weierstrass theorem (see
e.g.~\citep[Section~10.10]{Davidson-Donsig-2010}), on a compact set,
multivariate polynomials uniformly approximate any continuous function.
Intuitively speaking, we use a multivariate polynomial to approximate, on the
unit ball of $\R^d$, the indicator function corresponding to the intersection of
$m=K-1$ halfspaces. Within margin $\gamma$ along the decision boundary, we allow
the polynomial to attain arbitrary value. The polynomial separates examples in
one class from examples in the other classes. To be able to quantify $R'$,
$\gamma'$ and time complexities of evaluating $\phi(x)$ and $k(x,x')$, we need
to quantify certain parameters of the approximating polynomial. We construct two
different polynomials with different parameters. The parameters are quantified
in
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
stated below.

Before we state the theorems, recall that a polynomial of $d$ variables is a
function $p:\R^d \to \R$ of the form
$$
p(x) = p(x_1, x_2, \dots, x_d) = \sum_{\alpha_1, \alpha_2, \dots, \alpha_d} c_{\alpha_1, \alpha_2, \dots, \alpha_d} x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d}
$$
where the sum ranges over a finite set of $d$-tuples $(\alpha_1, \alpha_2,
\dots, \alpha_d)$ of non-negative integers and $c_{\alpha_1, \alpha_2, \dots,
\alpha_d}$ is a real coefficient. The \emph{degree} of a polynomial $p$, denoted
by $\deg(p)$, is the largest value of $\alpha_1 + \alpha_2 + \dots + \alpha_d$
for which the coefficient $c_{\alpha_1, \alpha_2, \dots, \alpha_d}$ is non-zero.
The \emph{norm of a polynomial} $p$ is defined as
$$
\norm{p} = \sqrt{\sum_{\alpha_1, \alpha_2, \dots, \alpha_d} \left(c_{\alpha_1, \alpha_2, \dots, \alpha_d} \right)^2 } \; .
$$
It is easy see that this is indeed a norm, since we can interpret it as a
Euclidean norm of the vector of the coefficients of the polynomial.

\begin{theorem}[Polynomial approximation of intersection of halfspaces I]
\label{theorem:polynomial-approximation-1}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$. There exists a
multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $p(x) \ge 1/2$ for all $\displaystyle x \in \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$
\item $p(x) \le -1/2$ for all $\displaystyle x \in \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$
\item $\displaystyle \deg(p) = \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil$
\item $\displaystyle \norm{p} \le \left( 188 \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil \right)^{\frac{1}{2} \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil}$
\end{enumerate}
\end{theorem}

>>>>

% Math symbols and commands
\newcommand{\calF}{\mathcal{F}}
\newcommand{\one}{\mathds{1}}
\newcommand{\R}{\mathbb{R}}  % set of real numbers
\newcommand{\indicator}[1]{\mathbf{1}\left[#1 \right]} % indicator
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle} % inner product
%\newcommand{\norm}[1]{\left\| #1 \right\|}  % norm of a vector or a matrix
%\DeclareMathOperator*{\deg}{deg}

\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\Exp}{\mathbf{E}}  % expected value
\DeclareMathOperator{\diff}{d \!} % differential
\DeclareMathOperator*{\polylog}{polylog}

@@

\input{introduction}

\input{related-work}

\input{notions-of-linear-separability}

\input{algorithm-for-strongly-linearly-separable-data}

\input{from-linear-separability-to-strong-linear-separability}

\input{proofs-of-polynomial-approximation-theorems}

\input{nearest-neighbor-algorithm}

\input{np-hardness-of-the-weak-labeling-problem}

\input{mistake-lower-bound-for-ignorant-algorithms}
